Fit by design: Developing seed-substrate combinations to adapt dike grasslands to microclimate

### 2023-07-03 manuscript.Rmd
Update manuscript.Rmd

---
title: "Fit by design: Developing substrate-specific seed mixtures for functional dike grasslands"
author: Markus Bauer*, Jakob K. Huber, Johannes Kollmann
output:
  officedown::rdocx_document:
    reference_docx: template_word.docx
csl: template_references_j_appl_ecol.csl
bibliography: bibliography.bib
---

@@

1.  Sowing is a well-established restoration technique to overcome dispersal limitation. Site-specific seed mixtures are most effective to achieve functional communities. This is especially important if the restored vegetation has to protect critical infrastructure like roadsides and dikes. Here, an improved seed--substrate combination will secure slope stability, make restorations more effective, and generate species-rich grasslands.
2.  A factorial field experiment addressed this topic on a dike at River Danube in SE Germany in 2018--2021. Within 288 plots, we tested three sand admixtures, two substrate depths, two seed densities and two seed mixture types (mesic hay meadow, semi-dry calcareous grassland) in north and south exposition, and measured the recovery completeness by calculating the successional distance to reference sites, the persistence of sown species, and the Favourable Conservation Status (FCS) of target species.
3.  Overall, the sown vegetation developed in the desired direction, but a recovery debt remained after four years, and some plots still showed similarities to negative references from ruderal sites. In north exposition, hay meadow-seed mixtures developed closer to the respective reference communities than dry-grassland mixtures.
4.  In south exposition, the sown communities developed poorly which might be due to a severe drought during establishment. This initial negative effect remained over the entire observation period.
5.  Sand admixture had a slightly positive effect on target variables, while the tested substrate depths, seed densities and mixture types had no effects on species persistence or FCS.
6.  Synthesis and applications: Site-adapted seed mixtures make restoration more effective. However, applying several seed--substrate combinations might foster beta diversity. Furthermore, additional management efforts are recommended, as they might be necessary to reduce the recovery debt, as well as re-sowing after unfavourable conditions. \clearpage \newpage

@@

Dikes are promising sites for the restoration of species-rich grasslands because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in agricultural landscapes [@Batori.2020]. Steep slopes (\>1:3) with different exposition, contrasting substrate layers and dense swards for erosion protection characterise these habitats [@Batori.2016; @berendse2015; @husicka2003]. Dikes can reconcile several ecosystem functions including both flood security and rich biodiversity [@teixeira2022], which can be fostered by an adapted seed--substrate combination.

@@

We used 288 plots of the size 2.0 m × 3.0 m, vertically oriented, halfway up the dike slopes, distributed over the north and south exposition, and arranged in six blocks (=replicates). The experiment used a split-plot design combined with a randomised complete block design (Figure \@ref(fig:map)). The split plot was created by the two expositions of the dike, where all 24 treatment combinations were tested, i.e., sand admixtures (0, 25, and 50%), soil depths (15 vs. 30 cm), the two seed mixture types, and two seed densities [4 vs. 8 g m^-2^, cf. @Kiehl.2010; @KleberLerchbaumer.2017].

@@

The ordination showed the species composition of seed mixtures and the development of the plots during four years (Figure \@ref(fig:nmds); 2D-stress: 0.21). The NMDS confirmed that the seed mixtures were variable, albeit distinctive for hay meadows and dry grasslands and confirming the intended direction of the vegetation development. As one exception, hay meadows in south exposition did not develop towards their seed-mixture compositions.

The reference sites had a larger variation than the seed mixtures and were close to the seed mixtures but hardly overlapped (Figure \@ref(fig:nmds)). The positions of the reference sites shifted to the left in comparison to the seed mixtures, which means in the direction of early-successional stages. Nonetheless, they still differed from the negative references of ruderal vegetation. Negative references were only available on the south exposition and they were located in the NMDS between the positive reference sites and the state of restored plots in 2021. Nevertheless, 33% of the 288 plots reached the state of the target habitat types by 2021 [EUNIS code R22, R1A, @Chytry.2020]. Hay meadow-seed mixtures led to a closer development to hay-meadow references than dry grasslands to their references (Figure \@ref(fig:effects)A, \@ref(fig:interactions)A). This was especially the case in north exposition (Figure \@ref(fig:nmds)).

@@

A statistically clear positive effectof the sand admixture was identified on the persistence of sown species and on the recovery rate, but no effects of seed density or substrate depth (Figure \@ref(fig:effects); Persistence: *R*<sup>2</sup><sub>m</sub> = 0.86, *R*<sup>2</sup><sub>c</sub> = 0.89; Recovery: *R*<sup>2</sup><sub>m</sub> = 0.90, *R*<sup>2</sup><sub>c</sub> = 0.92; FCS: *R*<sup>2</sup><sub>m</sub> = 0.81, *R*<sup>2</sup><sub>c</sub> = 0.85). The posterior distributions are also shown in the interaction plots that separate exposition and survey year (Figures \@ref(fig:interactions)). For all three response variables, the vegetation developed positively after one year, while the recovery rate slowed down in the following years. Both expositions revealed similar trends but for all responses, the values were clearly lower in south exposition, e.g., persistence values were on average more than 46% higher in north exposition (Figure \@ref(fig:interactions)B). The interactions of restoration treatments were neither clear nor strong. Persistence of both seed mixture types was slightly positively affected by sand admixture in north exposition (+ 6--7 ± 4%, Figure \@ref(fig:interactions)B). \clearpage \newpage

@@

The seed mixtures and their positive reference sites were similar but hardly overlapped (Figure \@ref(fig:nmds)). The position on the ordination suggests that the seed mixture represents a late-successional stage compared to the references. The NMDS shows a slightly better adaptation of hay meadows to the north exposition than of dry grasslands (Figure \@ref(fig:interactions)A). This can be expected from the requirements of hay meadows for mesic conditions, which can be provided on north-exposed dike slopes [@Oberdorfer.1993; @Batori.2020]. In southern exposition, the plots of hay meadows developed rather towards dry grassland references which indicates an ineffective restoration due to a non-adapted seed mixture.

The vegetation developed generally in the desired direction. The development after one year was very fast and afterwards very slow. For practitioners, it would be more cost-efficient to start monitoring not in the year of seeding because ruderal species are too dominant for a sensible interpretation of the results. After four years, the vegetation was still distinct from positive references and seed mixtures. In the south exposition, the plots were rather similar to the negative reference of dry ruderal vegetation typical of grassland restoration when perennial species are still developing to become dominant [@eckhoff2023]. The gap between goal and restoration outcome was also shown for other sowing experiments or restorations [@mitchley2012; @Engst.2016; @Kaulfuss.2022] or for dike vegetation compared with semi-natural reference grassland [@Batori.2016]. This result is not surprising since the 'recovery debt' is a general phenomenon of grassland restoration [@MorenoMateos.2017; @jones2018] and five years might be too short for the assembly of secondary grasslands [cf. @nerlekar2020]. The annual *B. hordeaceus*, seeded in autumn 2018 as nursery plant, decreased but was still present in 179 of 288 plots in 2021.

@@

Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but mixing with a mineral component need not necessarily improve the outcome [@chenot-lescure2022]. Similarly to @chenot-lescure2022, sand admixture reduced nutrient loads and led to higher persistence of sown species while an increase from 25% to 50% admixture did not further increase this effect. In addition, the effect only appeared in north exposition and the effect size of about 6% in the 4^th^ year of restoration was rather small. The Favourable Conservation Status (FCS) was hardly affected by the sand admixture, which corresponds to an experiment in a quarry [@chenot-lescure2022]. Substrate depth did not significantly affect persistence or FCS, similarly to earlier studies [@husicka2003; @baer2004]. Larger differences in soil depths might be necessary to observe negative effects by thicker substrate layers as was shown for prairies [@dornbush2010] or a thin substrate layer of \<15 cm, since most roots occur in the topsoil on dikes [@vannoppen2016]. Seed density had also no clear effect on persistence and FCS which fits the results of @Kaulfuss.2022, who found that a certain amount of seeds is necessary for a successful establishment of target species, but higher densities do not further improve the outcome, but rather have a slightly negative effect.

The vegetation in south exposition had a different species composition, which confirms the findings of @Batori.2016 in Hungary. However, the differences might also be due to methodical reasons, since the geotextile, which had been implemented on the southern slope, was removed after two weeks. This was unfortunate for at least some seedlings, and amplified by the intense drought in summer 2018 and 2019 [cf. @hari2020; @Larson.2021; @orrock2023]. The lasting negative effect on persistence and FCS on the southern slope suggests a legacy effect of adverse weather conditions after sowing as observed by other studies [@Stuble.2017; @Groves.2020]. These conditions during the establishment phase might have led to a special trajectory [@suding2004], and probably levelled the distinction of the seed mixture types in south exposition.

@@

Our aim was to identify perfect seed--substrate combinations regarding restoration effectiveness and biodiversity. For evaluating effectiveness, we measured the persistence of the sown species, and FCS for investigating plant biodiversity. However, we could not identify an interaction effect for neither one of these indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture. Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows vs dry grasslands). Although, both types are clearly phytosociologically and functionally distinct, they are still relatively close, because they contain shared species and develop under similar site conditions with modified sub-associations [Appendix A3, @husicka2003; @Oberdorfer.1993]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. However, these studies did not work with an experimental set up of different seed--substrate combinations, but analysed the result of habitat and biotic filtering after 1, 5 and 15yrs, respectively. Furthermore, the non-existence of ideal combinations could be explained by priority effects that means that the species of the imperfect-adapted seed mixture type could establish earlier and pre-empted the available niches for the species of related habitat types [@Fukami.2015].

@@

There were no perfect seed--substrate combinations and thus we conclude that a variation of seed mixture types and different substrates along restoration sections would promote biodiversity more than a single uniform solution [@bauer2023; @Holl.2022]. Negative effects of drought in the sowing season might require re-sowing. To close the recovery debt, the management adaptation might be promising since this is a crucial factor beside the restoration approach and the site characteristics [@Grman.2013; @tölgyesi2021]. For example, the introduction of sheep grazing on the experimental plots, which already exists in the surroundings, will modify the disturbance regime and improve dispersal. Overall, our results support the finding that restored dike grasslands can promote biodiversity in agricultural landscapes [@Batori.2020]. However, the recovery debt highlights the fact that restored grasslands cannot substitute old-growth grasslands [@nerlekar2020].

>>>>

---
title: "Fit by design: Developing seed-substrate combinations to adapt dike grasslands to microclimate"
author: Markus Bauer*, Jakob K. Huber, Johannes Kollmann
output:
  officedown::rdocx_document:
    reference_docx: template_word.docx
csl: template_references_j_appl_ecol.csl
bibliography: bibliography.bib
---

@@

1.  Sowing is a well-established restoration technique to overcome dispersal limitation. Seed mixtures adapted to certain environmental conditions like substrate or microclimate are most effective to achieve functional communities. This is especially important if the restored vegetation has to protect critical infrastructure like roadsides and dikes. Here, an improved seed--substrate combination will secure slope stability, make restorations more effective, and generate species-rich grasslands.
2.  A factorial field experiment addressed this topic on a dike at River Danube in SE Germany in 2018--2021. Within 288 plots, we tested three sand admixtures, two substrate depths, two seed densities and two seed mixture types (mesic hay meadow, semi-dry calcareous grassland) in north and south exposition, and measured the recovery completeness by calculating the successional distance to reference sites, the persistence of sown species, and the Favourable Conservation Status (FCS) of target species.
3.  Overall, the sown vegetation developed in the desired direction, but a recovery debt remained after four years, and some plots still showed similarities to negative references from ruderal sites. In north exposition, hay meadow-seed mixtures developed closer to the respective reference communities than dry-grassland mixtures.
4.  In south exposition, the sown communities developed poorly which might be due to a severe drought during establishment. This initial negative effect remained over the entire observation period.
5.  Sand admixture had a slightly positive effect on target variables, while the tested substrate depths, seed densities and mixture types had no effects on species persistence or FCS.
6.  Synthesis and applications: Site-adapted seed mixtures make restoration more effective. However, applying several seed--substrate combinations might foster beta diversity. Furthermore, additional management efforts are recommended, as they might be necessary to reduce the recovery debt, as well as re-sowing after unfavourable conditions. \clearpage \newpage

@@

Dikes are promising sites for the restoration of species-rich grasslands because they can increase habitat area and connectivity of semi-natural grasslands and therefore significantly contribute to biodiversity conservation in agricultural landscapes [@Batori.2020]. In addition, dikes are interesting for experiments with seed-substrate combinations because the combinations can be tested on contrasting microclimates of the steep slopes (\>1:3) with different expositions.

contrasting substrate layers and dense swards for erosion protection characterise these habitats [@Batori.2016; @berendse2015; @husicka2003]. Dikes can reconcile several ecosystem functions including both flood security and rich biodiversity [@teixeira2022], which can be fostered by an adapted seed--substrate combination.

@@

We used 288 plots of the size 2.0 m × 3.0 m, vertically oriented, halfway up the dike slopes (1:2), distributed over the north and south exposition, and arranged in six blocks (=replicates). The experiment used a split-plot design combined with a randomised complete block design (Figure \@ref(fig:map)). The split plot was created by the two expositions of the dike, where all 24 treatment combinations were tested, i.e., sand admixtures (0, 25, and 50%), soil depths (15 vs. 30 cm), the two seed mixture types, and two seed densities (4 vs. 8 g m^-2^). @Kiehl.2010 recommend 1--5 g m^-2^ for grassland restoration and @KleberLerchbaumer.2017 recommend an increased density of 5--8 g m^-2^ for slopes.

@@

The ordination showed the species composition of seed mixtures and the development of the plots during four years (Figure \@ref(fig:nmds); 2D-stress: 0.21). The NMDS confirmed that the seed mixtures were variable, albeit distinctive for hay meadows and dry grasslands and confirming the intended direction of the vegetation development. As one exception, hay meadows in south exposition did not develop towards their seed-mixture compositions as expected in hypothesis H1.

The reference sites had a larger variation than the seed mixtures and were close to the seed mixtures but hardly overlapped (Figure \@ref(fig:nmds)). The positions of the reference sites shifted to the left in comparison to the seed mixtures, which means in the direction of early-successional stages. Nonetheless, they still differed from the negative references of ruderal vegetation. Negative references were only available on southern slopes and they were located in the NMDS between the positive reference sites and the state of restored plots in 2021. Nevertheless, 33% of the 288 plots reached the state of the target habitat types by 2021 [EUNIS code R22, R1A, @Chytry.2020]. Hay meadow-seed mixtures led to a closer development to hay-meadow references than dry grasslands to their references (Figure \@ref(fig:effects)A, \@ref(fig:interactions)A). This was especially the case in north exposition (H1; Figure \@ref(fig:nmds)).

@@

A statistically clear positive effect of the sand admixture (H2) was identified on the persistence of sown species and on the recovery rate, but no effects by substrate depth (H2) or seed density (H3) (Figure \@ref(fig:effects); Persistence: *R*<sup>2</sup><sub>m</sub> = 0.86, *R*<sup>2</sup><sub>c</sub> = 0.89; Recovery: *R*<sup>2</sup><sub>m</sub> = 0.90, *R*<sup>2</sup><sub>c</sub> = 0.92; FCS: *R*<sup>2</sup><sub>m</sub> = 0.81, *R*<sup>2</sup><sub>c</sub> = 0.85). The posterior distributions are also shown in the interaction plots that separate exposition and survey year (Figures \@ref(fig:interactions)). For all three response variables, the vegetation developed positively after one year, while the recovery rate slowed down in the following years. Both expositions revealed similar trends but for all responses, the values were clearly lower in south exposition, e.g., persistence values were on average more than 46% higher in north exposition (Figure \@ref(fig:interactions)B). The interactions of restoration treatments were neither clear nor strong (H2, H3). Persistence of both seed mixture types was slightly positively affected by sand admixture in north exposition (H2; + 6--7 ± 4%, Figure \@ref(fig:interactions)B). \clearpage \newpage

@@

The seed mixtures and their positive reference sites were similar but hardly overlapped (Figure \@ref(fig:nmds)). The position on the ordination suggests that the seed mixture represents a late-successional stage compared to the references. The NMDS shows a slightly better adaptation of hay meadows to the north exposition than of dry grasslands what we had more clearly expected in H1 (Figure \@ref(fig:interactions)A). This can be expected from the requirements of hay meadows for mesic conditions, which can be provided on north-exposed dike slopes [@Oberdorfer.1993; @Batori.2020]. In south exposition, the hay meadow plots developed rather towards dry grassland references which indicates an ineffective restoration due to a not adapted seed mixture to the microclimatic conditions of southern slopes.

The vegetation developed generally in the desired direction. The development after one year was very fast and afterwards very slow. For practitioners, it would be more cost-efficient to start monitoring not in the year of seeding because ruderal species are too dominant for a sensible interpretation of the results. After four years, the vegetation was still distinct from positive references and seed mixtures. In the south exposition, the plots were rather similar to the negative reference of dry ruderal vegetation typical of grassland restoration when perennial species are still developing to become dominant [@eckhoff2023]. The gap between goal and restoration outcome was also shown for other sowing experiments or restorations [@mitchley2012; @Engst.2016; @Kaulfuss.2022] or for dike vegetation compared with semi-natural reference grassland [@Batori.2016]. This result is not surprising since the 'recovery debt' is a general phenomenon of grassland restoration [@MorenoMateos.2017; @jones2018] and five years might be too short for the assembly of secondary grasslands [cf. @nerlekar2020]. The annual *B. hordeaceus*, seeded in autumn 2018 as nursery plant, decreased but was still present in 179 of 288 plots in 2021. Reasons for the recovery debt might not only be abiotic conditions but can also be biotic factors like missing mycorrhiza in the substrates [@koziol2017] or the post-restoration management needs to be developed [@tölgyesi2022].

@@

Restoration on agricultural soils can have limited success due to high nutrient loads [@walker2004] but mixing with a mineral component need not necessarily improve the outcome [@chenot-lescure2022]. Similarly to @chenot-lescure2022, sand admixture reduced nutrient loads and led to higher persistence of sown species as expected by H2 while an increase from 25% to 50% admixture did not further increase this effect. In addition, the effect only appeared in north exposition and the effect size of about 6% in the 4^th^ year of restoration was rather small. The Favourable Conservation Status (FCS) was hardly affected by the sand admixture, which corresponds to an experiment in a quarry [@chenot-lescure2022] but not to our expectations (H2). An increase of substrate depth from 15 to 30 cm did neither significantly affect persistence or FCS, similarly to earlier studies [@husicka2003; @baer2004]. Larger differences in soil depths might be necessary to observe negative effects by thicker substrate layers as was shown for prairies [@dornbush2010] or a thin substrate layer of \<15 cm, since most roots occur in the topsoil on dikes [@vannoppen2016]. Seed density had also no clear effect on persistence and FCS which is contrary to H3 but fits the results of @Kaulfuss.2022, who found that a certain amount of seeds is necessary for a successful establishment of target species, but higher densities do not further improve the outcome, but rather have a slightly negative effect.

The vegetation in south exposition had a more ruderal and xerophytic species compositions than in north exposition, which contrasts @Batori.2016 who found different compositions in south exposition only for riverside slopes which caused a more mesotrophic vegetation. The differences in our experiment might be due to methodical reasons, since the geotextile, which had been implemented on the southern slope, was removed after two weeks. This was unfortunate for at least some seedlings, and amplified by the intense drought in summer 2018 and 2019 [cf. @hari2020; @Larson.2021; @orrock2023]. The lasting negative effect on persistence and FCS on the southern slope suggests a legacy effect of adverse weather conditions after sowing as observed by other studies [@Groves.2020; @atkinson2023]. These conditions during the establishment phase might have led to a special trajectory [@suding2004], and probably leveled the distinction of the seed mixture types in south exposition.

@@

Our aim was to identify perfect seed--substrate combinations regarding restoration effectiveness and biodiversity (H2, H3). For evaluating effectiveness, we measured the persistence of the sown species, and FCS for investigating plant biodiversity. However, we could not identify an interaction effect for neither one of these indices. We would have expected a better performance of hay-meadow seed mixtures with lower sand admixture and for dry grasslands with higher sand admixture (H2). Our results suggest that, at least after four years, the substrate conditions are within the range of both seed mixture types (hay meadows vs dry grasslands). Although, both types are clearly phytosociologically and functionally distinct, they are still relatively close, because they contain shared species and develop under similar site conditions with modified sub-associations [Appendix A3, @husicka2003; @Oberdorfer.1993]. Other grassland studies could identify more or less clear interactions of opposing habitat preferences or functional traits along the gradients of productivity, moisture and nutrients [@Zirbel.2020; @Freitag.2021; @Kaulfuss.2022]. However, these studies did not work with an experimental set up of different seed--substrate combinations, but analysed the result of habitat and biotic filtering after 1, 5 and 15yrs, respectively. Furthermore, the non-existence of ideal combinations could be explained by priority effects that means that the species of the imperfect-adapted seed mixture type could establish earlier and pre-empted the available niches for the species of related habitat types [@Fukami.2015].

@@

There were no perfect seed--substrate combinations and thus we conclude that a variation of seed mixture types and different substrates along restoration sections would promote biodiversity more than a single uniform solution [@bauer2023; @Holl.2022]. Negative effects of drought in the sowing season might require re-sowing. Restoration projects might account for the increasing frequency of droughts [@naumann2018] by include a re-sowing in planning or by combining seeding with hay transfer to improve the microclimate during establishment [@eckstein2005]. We expect a minor effect of succession in the next ten years, which requires further interventions to close the recovery debt. The management adaptation might be promising since this is a crucial factor beside the restoration approach and the site characteristics [@Grman.2013; @tölgyesi2022]. For example, the introduction of sheep grazing on the experimental plots, which already exists in the surroundings, will modify the disturbance regime and improve dispersal. Overall, our results support the finding that restored dike grasslands can promote biodiversity in agricultural landscapes [@Batori.2020]. However, the recovery debt highlights the fact that restored grasslands cannot substitute old-growth grasslands [@nerlekar2020].
